---
title: "Non-uniform Berry Essen Bounds"
date: "`r Sys.Date()`"
output: html_document
bibliography: references.bib
execute:
  echo: false
warning: false
message: false
error: false
editor:
  markdown:
    wrap: 72
---

Twice in the last week, I've had computer scientists laugh about the
insanity of asymptotics in Statistics[^1]. The insanity comes from the
assumption that is never true: $n = \infty$. As someone working on
asymptotic approximations in Statistics, I feel the need to defend its
usefulness by showing how people have justified it in the past, but also
think through ways to improve upon the justification for using normal
models.

[^1]: I remark that computer scientists are apparently very against
    normal approximations and enjoy their finite sample correct
    concentration inequalities. This tracks with my [earlier post on
    asymptotics](Asymptotics.html).

## $\Delta$ isn't Zero

In Statistics, we often assume $n = \infty$ and this makes people feel
uncomfortable. Outside of statistics, engineers linearize systems
assuming that $\Delta = 0$ in the Taylor expansion. If you have some
complex function $f(x)$, let's just analyze the second order expansion.
PID controllers, for instance, are second order controllers and are
incredibly popular.

The assumptions are essentially the same. Assuming $\Delta = 0$ is the
same as assuming $n = \infty$. The proof I learned of the CLT [@tao2015clt] is
the Lindeberg swapping trick. You take two derivatives (hence the need
for two finite moments) and write all higher order terms as $o(1)$. In
some regard the Gaussian is the first order random variable
approximation.

There is interesting psychology happening that brings about complete
terror when you assume $n = \infty$ but does not bring it out when
assuming $\Delta = 0$. I suspect humans often see numbers near $0$ in
daily life, but will never see a number near infinity. Moreover, we can
imagine an infinitely small amount of time, but can never imagine an
infinitely large amount of time.

## Berry-Essen Bounds

If $n \neq \infty$ then how can we ever use a normal approximation? In
other words, what is the approximation error? The answer to this
question is the very classical result in probability called the
Berry-Essen bound.

### Recall the CLT

Recall the statement of the CLT: Let $X_1, \dots \sim \mathsf P$ be
i.i.d with variance $\sigma^2 < \infty$ and zero mean. Define
$S_n = \sum_{i=1}^{n} X_i$ to be the partial
sum. Then $$
\lim_{n \to \infty} \left|\mathsf P\left(\frac{S_n}{\sigma\sqrt{n}} \leq z_{\alpha}\right) - \alpha \right| = 0,
$$ where $z_\alpha$ is the $\alpha$ quantile of a standard normal. That
is $\mathsf P(Z < z_\alpha) = \alpha$ where $Z \sim \mathcal N(0,1)$.
The theorem says if you look at large enough $n$ your scaled partial
sums are essentially Gaussian.

### Approximation Error

The obvious question is what is "large enough". In my first STAT class
at Duke, I had a z-score table and a t-table. I was told that if
$n > 30$ then we can use a z-test. 

To motivate the approximation error, assume you $20$ Bernoulli
datapoints: $X_1, \dots, X_{20}$. You want to perform a hypothesis test
asking if $p = 0.5$ or if $p > 0.5$. You have not learned about
Neyman-Pearson and the likelihood ratio test, so you instead compute a
p-value using the Gaussian approximation. That is you write $$
T_{20} = \frac{\sum_{i=1}^{20} (X_i - 0.5)}{\sqrt{20 \times 0.25}} \overset{H_0}{\approx } \mathcal N(0, 1).
$$ You observe $15$ ones and $5$ zeroes, so you can calculate $T_{20}$ to
be

```{r}
x <- sample(c(rep(1, 15), rep(0, 5)))
print(x)
T_20 <- (sum(x - 0.5))/(sqrt(20 * 0.25))
print(paste0("T_20: ", round(T_20, 4)))
```

and you can find the p-value by looking at the tail of the normal
distribution:

```{r}
p_val <- pnorm(T_20, lower.tail = F)
print(paste0("p-val: ", round(p_val, 3)))
```

Great! The p-value less than 0.05 so we can reject![^2] However, we
remember that this is an approximate p-value that becomes exact when
$n = \infty$. But $n = 20$, which I would claim is *much* smaller than
$\infty$. Should we be worried? What is the approximation error?

[^2]: Unfortunately I've asked friends outside of statistics what a
    p-value is and they've responded that it needs to be less than 0.05.

### Berry-Essen Bound

What we are asking is how close is $T_{20}$ to a standard normal random
variable. The Berry-Esseen bound [@berry1941; @esseen1942] gives us an answer. It states that if
you assume a finite third moment $\mathbb{E}(|X_1 - \mathbb E(X_1)|^3) = \rho < \infty$,
then you can get a uniform error band. $$
\sup_{x \in \mathbb R}\left | \mathsf P\left(\frac{S_n}{\sigma\sqrt{n}} \leq x\right) - \mathsf P\left(Z \leq x\right) \right | \leq \frac{C \rho}{\sigma^3\sqrt{n}},
$$ where $C$ is some universal constant around $0.5$. Basically, the
theorem says that if you take the scaled partial sums
$\frac{S_n}{\sigma\sqrt{n}}$ then you expect them to be Gaussian if $n$ is
large. How Gaussian? The distance from a true Gaussian $Z$ decreases at
a $\frac{1}{\sqrt n}$ rate and depends on the variance and third moment.

**Return to the Binomial Example**

We found that the p-value is $0.013$ and define $z^\star$ to be the
normal quantile at this p-value. We want to know how big the following
quantity is:

$$
\left | \mathsf P\left (\frac{S_n}{\sigma\sqrt{n}} > z^\star \right) - \mathsf P\left (Z > z^\star \right) \right |.
$$

We're reporting the number on the right side of the absolute value but
what we really care about is the number on the left side of the absolute
value. Recall the statement of the Berry-Essen bound. It says for any
quantile value $x \in \mathbb R$ the distance from the true probability
and the approximate one is upper bounded at a $1/\sqrt{n}$ rate. 

The statement is true for any $x \in \mathbb R$, so it ought to be true
for $z^\star$ since $z^\star$ is in $\mathbb R$. You can calculate that $\rho/\sigma^3$ is 1 since this ratio captures the asymmetry of random variables. Taking $C \approx 0.47$ gives the maximum error is 0.105.

Therefore
$$
\left | \mathsf P\left (\frac{S_n}{\sigma\sqrt{n}} > z^\star \right) - \mathsf P\left (Z > z^\star \right) \right | \leq 0.1.
$$
and your p-value $0.01$ which was before less than 0.05 could now be at most $0.11$,
which is certainly greater than 0.05!

## Uniformity is Overkill

The usual Berry-Essen bound gives a uniform deviance from the 
true partial sums with a Gaussian random variable. We used the following logic:
The Berry-Essen bound is true for any $x \in \mathbb R$ so it ought to be 
true for $z^\star$. This jump loses a lot of information and hence
the bound becomes quite a bit more loose. 

The intuition is that the tails of a distribution function are a lot less variable
than the middle of a distribution function [^3]. I've wondered
about this gap of the Berry-Esseen bound this year and finally took time
to search the literature on what are called
Non-Uniform Berry-Esseen Bounds [@nagaev1965]. They are *strictly* stronger than the uniform counterparts because they imply the uniform counterparts.

The non-uniform bound assumes the exact same set up but now we have
$$
\left | \mathsf P\left(\frac{S_n}{\sigma\sqrt{n}} \leq x\right) - \mathsf P\left(Z \leq x\right) \right | \leq \frac{C' \rho}{\sigma^3\sqrt{n}}\left(\frac{1}{1 + |x|^3}\right).
$$
This formulation has a correction factor tied to the original rate. Notice that if $x$ is very extreme then the fraction is $< 1$ and thus shrinks the distance. Moreover, this statement implies the uniform one since taking $\sup x$ on both sides reduces the correction factor to just one.

## The Constant C'
The tight constant for the uniform bound is a bit less than $0.5$. And for the
non-uniform bound? $30$! See @pinelis2013, Section 3.1, for an entertaining historical account of this constant, which includes some algebraic mistakes along the way.

If we were to apply the non-uniform bound in our Bernoulli example, we get 
a distance of $0.4$ whereas the uniform bound gives us a max distance of $0.1$. We're in a worse situation using the non-uniform bound because of the constant.
Intuitively we're ok using CLT-based hypothesis testing because it is 
approximately correct. The uniform Berry-Essen is too conservative 
and the non-uniform version has too large of a constant that one
would expect could be tightened but the tools aren't there.




[^3]: This is because quantiles are binomial random variables and more extreme $p$ has smaller variance.

