---
title: "Honest Hypothesis Tests"
format: html
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

There are many types of hypothesis tests. One such type is the "honest"
hypothesis test. This description may conjure an image in your mind.
What are "honest" hypothesis tests compared to their regular
counterparts? Maybe a hypothesis test is honest if it does not double
dip, have some stopping rule, or has well founded assumptions. Honest
can mean many things, but its description today will be none of these
things. Instead, the modifier "honest" hypothesis tests stems from
@li1989. Its meaning is much more esoteric. It can only be explained to
someone who has taken a graduate level analysis class.

### Preliminaries

We'll begin with @bahadur1956 which stated an impossibility result. Let
$X_1, X_2, \ldots X_n$ be i.i.d. with mean $\mu = \mu_p$ and variance
$\sigma^2(P) < \infty$. We do not want to put any other restrictions on
the random variables $(X_n)$ when we do statistical inference. So we
consider a broad class of distributions $\mathcal{P}$, whose restriction
is that random variables have finite variance. Our null hypothesis will
be $\mathcal{P}_0 \subset \mathcal{P}$ where
$\mathcal{P}_0 \equiv \{P : \mu = \mu_0\}$ for some fixed $\mu_0$. We
don't have any assumptions on $(X_n)$, knowledge of their likelihoods.
Perhaps the only thing we can possibly do is appeal to asymptotics. We
have a finite second moment after all... We recall the central limit
theorem (CLT), which says that $\forall P \in \mathcal{P}$,

$$
\frac{1}{\sqrt{n}}\left(S_n - n\mu_p\right) \Rightarrow N(0, \sigma^2).
$$

We hope $n$ is large enough that that the LHS is approximately the RHS.
We still don't know the variance $\sigma^2$ which indicates our use of a
t-test.[^1] We'll define our test $\phi_n$ to be

[^1]: If $n$ is large $t,z$ will give you the same answer anyway

$$
\phi_n(X_1, \ldots, X_n) = \mathbf{1}\left\{\sqrt{n}\bar X_n > \hat \sigma_n t_{1-\alpha, n-1}\right\}.
$$

So far so good. We can write down the power function of $\phi_n$ which
is

$$
\mathbb{E}_{\mathsf P}(\phi_n) \leq \alpha
$$and by the CLT we know that

$$
\limsup_{n \to \infty} \mathbb{E}_{\mathsf P}(\phi_n) \leq \alpha.
$$

We can check this empirically, for instance, when
$Z_n \sim {\rm Exp}(1) -1$.

```{r}
max_n <- 8e2
ns <- seq(10, max_n, by = 15)

res <- sapply(ns, \(n) {
  mean(replicate(1e4, {
    z <- rexp(n) - 1
    sqrt(n) * mean(z) / sd(z) > qt(0.95, n - 1)
  }))
})

plot(ns, res, type = "l",
     xlab = "n", ylab = "Rejection rate",
     main = "Empirical size of t-test (one-sided, exp-1)")
abline(h = 0.05, col = "red", lty = 2)
```

While the exponential distribution is a nice $\mathsf P$ is in our large
class $\mathcal{P}$, so is other more adversarial $\mathsf P$. Let
$\mathsf P_n$ be the distribution defined as follows

$$
\mathsf P_n(X) = \begin{cases} n & 1/n^2 \\ -n & 1/n^2 \\ 0 & {\rm o.w.}\end{cases}.
$$

We can check that the mean of this distribution is zero and it has a
finite second moment. Let's calculate the power function for any
$\mathsf P_n$ in this sequence. I claim the following still holds

$$
\sup_{n \in \mathbb{N}} \limsup_{n \to \infty} \mathbb{E}_{\mathsf P_n}(\phi_n) \leq \alpha,
$$so nothing is wrong here. This statement is true because of the CLT.
For any fixed $n$ the distribution is will converge. However, if we
change $\lim$ and $\sup$ we cannot guarantee this. Instead we have

$$
\limsup_{n \to \infty} \sup_{n \in \mathbb{N}} \mathbb{E}_{\mathsf P_n}(\phi_n) > \alpha
$$

In addition, each $\mathsf P_n$ has a finite \$2 + \delta\$ moment for
any $\delta > 0$. The first two facts are true for any $n$ and thus
$\mathsf P_n \in \mathcal{P},\forall n$. However, the last fact is not
true uniformly. Namely

$$
\sup_{\mathsf P \in \mathcal{P}} \mathbb{E}_{\mathsf P}(X^{2 + \delta}) = \infty.
$$

I was taught the assumption above is called the "Crystal Ball Condition"
@resnick1999. It is a sufficient but not necessary condition that is
usually the easiest to check.

I'll write the CLT we stated at the beginning again replacing
$\forall P \in \mathcal{P}$ with a $\sup_{P \in \mathcal{P}}$ and the
$\Rightarrow$ as a $\lim_{n \to \infty}$. The *pointwise* CLT more
explicitly claims

$$
\sup_{P \in \mathcal{P}} \lim_{n \to \infty} \sup_{x \in \mathbb{R}} |\Pr\left(\frac{1}{\sqrt{n}}(S_n-n\mu_p) \leq x\right) - \Pr(Z \leq x)| = 0.
$$

The **main use of uniformly "honest" tests** **is the ability to swap**
$\sup$ and $\lim$ . The issue we ran into above is the issue of
uniformity. In a measure theory class you may run into issues of uniform
integrability (UI). This topic and UI are intricately linked. Thus for a
class of distributions, I'll discuss later we can discuss the *uniform*
CLT

### Does this matter in practice?

I want to think through the question: is this either 1) a fun
mathematical exercise for statisticians to work on or 2) is it something
to add (perhaps near the top) to the million other things a statistician
should check when doing a data analysis.

At first I thought it was option 1. I thought this is just a fun fact
about hypothesis tests. Nobody doing a confidence interval in R should
worry about this messing up their analysis. But it is actually a third
that combines 1) and 2). I'll try to convince you and also convince
myself the use of getting trained in theoretical statistics.

### Yes & No

The answer to "does this matter" is option 3. No, it doesn't matter and
also it does *really* matter. A very statistician response. I'll first
explain why I think it doesn't matter.

\
As the analyst we're assuming finite mean and variance. The *entire*
problem vanishes if you assume that

for any $\delta > 0$, the \$2 + \delta\$'th moment exists. So if someone
in the audience is upset about uniformity, just tell them you assumed an
incrementally larger moment exists. The entire problem disappears and
you can go on t-testing away. This was my thinking before I learned more
about the problem. The tiniest change in assumption makes the problem go
away and its an assumption nobody will fight you on. So why do I think
it is something to think about when doing statistics?

When we talk about $\sup$ and $\lim$ and berry-essen bounds, the
problems become very abstract. However, what is real when you're staring
at a dataset and you think about the next observation coming in is how
sensitive is your analysis to tails? The problem with tails is you don't
know where they are until they are there. Take our adversarial heavy
tail example. We can simulate data from this process, which we do below.

```{r}
tail_sims <- replicate(1e3,{
  n <- 1e4
  x <- rbinom(1e5, size = 1, 1/n^2)
  which(x == 1)[1]
}
)
hist(tail_sims)
```

Maybe after 10,000 days using AI, we can be confident that chatbots are
not an adversary. But you never know! Bad events can be hiding in the
tails of the distribution. Even if the distribution is behaved enough to
allow a CLT.

A good example of tail events causing havoc is the use of the Gaussian
copula in the 2008 financial crisis @mackenzie2014.
