---
title: "Sequential Testing for Undergrads: Experimental Design"
format:
  html:
    toc: true
    code-fold: false
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

## Prerequisites

Assume knowledge of:

1.  Hypothesis testing, $p$-values
2.  Causal inference (ATE)
3.  Likelihoods, LRTs

------------------------------------------------------------------------

## The Problem with Peeking

What's wrong with \[insert bad stopping time\]? R simulations show
under-coverage. $\tau$ (confusing since $\tau$ is also ATE) is a
stopping time.

$X_\tau$ is random because $X_n$ is random **and** $\tau$ is random. The
core issue:

$$P\!\left(X_\tau > z_{1-\alpha/2}\right) \neq P\!\left(X_n \geq z_{1-\alpha/2}\right)$$

### What to do?

1.  Power calculations $\Rightarrow$ fix sample size beforehand
2.  Just guess.

------------------------------------------------------------------------

## Simulation: Repeated $z$-Tests Under $H_0$

**Setup.** Under $H_0$, draw $X_i \overset{\text{iid}}{\sim} N(0,1)$ and
compute a running $z$-test after each new observation. Stop and reject
as soon as $|Z_t| > z_{0.975} = 1.96$. Even though $H_0$ is true, this
"peek-and-stop" strategy rejects far more than 5% of the time.

```{r}
set.seed(42)

n_sim  <- 10000
n_max  <- 200
alpha  <- 0.05
z_crit <- qnorm(1 - alpha / 2)

# Peek after every observation; stop if |Z_t| > 1.96
peek_and_stop <- function(n_max, z_crit) {
  x <- rnorm(n_max)
  for (t in 2:n_max) {
    z_t <- mean(x[1:t]) * sqrt(t)
    if (abs(z_t) > z_crit) return(TRUE)
  }
  return(FALSE)
}

# Correct fixed-n test
fixed_n_test <- function(n_max, z_crit) {
  x <- rnorm(n_max)
  abs(mean(x) * sqrt(n_max)) > z_crit
}

reject_peek  <- mean(replicate(n_sim, peek_and_stop(n_max, z_crit)))
reject_fixed <- mean(replicate(n_sim, fixed_n_test(n_max, z_crit)))

cat(sprintf("Nominal alpha          : %.3f\n", alpha))
cat(sprintf("Fixed-n rejection rate : %.3f\n", reject_fixed))
cat(sprintf("Peek-and-stop rate     : %.3f\n", reject_peek))

round_reject_peek <- round(reject_peek * 100)
```

The peek-and-stop strategy rejects $H_0$ roughly `r round_reject_peek`%
of the time despite $\alpha = 0.05$ — a large inflation of Type I error.
This gets *worse* as `n_max` grows: given infinite peeks you will
eventually reject with probability 1 under the null.

```{r}
#| fig-cap: "Type I error explodes as you allow more peeks"

library(ggplot2)

peek_counts <- c(10, 25, 50, 100, 200, 500, 1000)

type1 <- sapply(peek_counts, function(n) {
  mean(replicate(n_sim, peek_and_stop(n, z_crit)))
})

ggplot(data.frame(peeks = peek_counts, type1 = type1), aes(peeks, type1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  annotate("text", x = 50, y = 0.07, label = "Nominal alpha = 0.05",
           color = "red", size = 3.5) +
  scale_x_log10() +
  labs(x = "Max number of peeks (log scale)",
       y = "Empirical Type I error",
       title = "Peek-and-stop inflates Type I error") +
  theme_minimal()
```

------------------------------------------------------------------------

## The Fix: Sequential Testing

First, why should we do sequential testing? Why not just prespecify $n$?
There is a theorem, which is hard to write down exactly, but I call it
the Always Do Sequential Testing Theorem @wald1945. It states
approximately that if you pick $n$ via a power calculation, on average
you could have gotten away with *fewer* samples had you used a
sequential test instead.

### What is a Sequential Test?

1.  **SPRT** — Sequential Probability Ratio Test
2.  **Asymp CS** — Asymptotic Confidence Sequences

------------------------------------------------------------------------

## SPRT

From Wald (1945). Given distributions $P$ and $Q$, test $P$ vs. $Q$.

$$X_1, \ldots, X_n \overset{\text{iid}}{\sim} P \text{ or } Q?$$

**Example:** $P \sim N(0,1)$, $Q \sim N(\theta, 1)$, $\theta > 0$.

The LRT statistic is $\frac{q_\theta}{p_\theta}(X_1, \ldots, X_n)$. That
is the ratios of densities. You can reject when this is "large" and you
calibrate under $P$. The SPRT extends this to sequential data
$X_1, X_2, \ldots$

**Specify:** $\alpha$ (Type I error, e.g. 0.05), $\beta$ (power, e.g.
0.8).

Platforms like [Statsig](https://www.statsig.com/) implement this and it
is quite easy to use.

**Problem:** Need to know likelihoods!

------------------------------------------------------------------------

## SPRT: Simple Example

**Setting.** $P = N(0,2)$, $Q = N(1, 2)$ (i.e. $\theta = 1$). The
log-likelihood ratio after $n$ observations is:

$$\Lambda_n = \log \frac{dQ}{dP}(X_1,\ldots,X_n)
= \sum_{i=1}^n \left[ X_i - \tfrac{1}{2} \right]$$

Wald's thresholds: reject $H_0$ ($P$) when
$\Lambda_n \geq \log(1/\alpha)$, accept $H_0$ when
$\Lambda_n \leq \log(\beta)$.

```{r}
set.seed(123)
#install.packages(SPRT)
library(SPRT)
sd <- 5
sprt_normal <- function(data){
  sprt(data, alpha = 0.05, beta = 0.05, p0 = 0, 
       p1 = 1, dist = "normal", sigma = sd)
}

# --- Under H0: X ~ N(0,1) ---
cat("=== Under H0: X ~ N(0,1) ===\n")
h0_results <- replicate(n_sim, {
  x <- rnorm(500, mean = 0, sd)
  sprt_normal(x)$decision
})
cat(sprintf("Reject H0 : %.3f  (should be <~%.2f)\n",
            mean(h0_results == "Reject H0"), alpha))

# --- Under H1: X ~ N(1,1) ---
cat("=== Under H1: X ~ N(1,1) ===\n")
h1_results <- replicate(n_sim, {
  x <- rnorm(500, mean = 1, sd)
  sprt_normal(x)$decision
})
cat(sprintf("Reject H0 : %.3f  (power, should be >~%.2f)\n",
            mean(h1_results == "Reject H0"), 1 - 0.05))
```

```{r}
#| fig-cap: "Single SPRT path under H1. The LLR crosses the rejection boundary early."

# lets plot if x is from the alternative
x <- rnorm(500, mean = 1, sd)
sprt_plot(sprt_normal(x))
```

### SPRT vs. Fixed-$n$: Average Sample Size

One key payoff of the SPRT is that it stops *early* on average.

```{r}
set.seed(99)

sample_sizes <- function(dist = c("H0", "H1"), n_sim = 5000) {
  dist <- match.arg(dist)
  replicate(n_sim, {
    x <- if (dist == "H0") rnorm(2000) else rnorm(2000, mean = 1)
    res <- sprt_normal(x)
    res$n
  })
}

ss_h0 <- sample_sizes("H0")
ss_h1 <- sample_sizes("H1")

cat(sprintf("Mean sample size under H0 : %.1f\n", mean(ss_h0)))
cat(sprintf("Mean sample size under H1 : %.1f\n", mean(ss_h1)))

# you can calculate this with some algebra.
fixed_n_required <- function(mu1, sigma, alpha = 0.05, beta = 0.05) {
  ceiling(((qnorm(1 - alpha) + qnorm(1 - beta)) * sigma / mu1)^2)
}
cat(sprintf("Fixed-n required (mu1=%.1f, sigma=%.1f, alpha=%.2f, power=%.0f%%): ~%d\n",
            1, sd, 0.05, (1 - 0.05) * 100,
            fixed_n_required(1, sd, 0.05, 0.05)))
```

The SPRT typically stops well before the fixed-$n$ sample size — this is
the Wald-Wolfowitz theorem in action.

## Testing to Confidence Sequences

Lastly, we'll talk about moving from hypothesis testing to confidence
intervals. There are a few things that aren't very nice about the SPRT.
It requires the analyst to

1.  Know the distribution (and variance)
2.  Specify the null and the alternative

In Statistics there is a dual between testing and confidence intervals.
We showed before that one cannot do repeated z-tests. It might seem
obvious then that one cannot do repeated confidence intervals.
Therefore, we'll introduce work on ***confidence sequences.*** Work on
confidence sequences is very new! In this work we'll look at a method
from @waudby-smith2024 with an application for tracking the ATE in an
experiment.

We consider the following simple experiment where `n` is our number of
total samples,

`p_0,p_1` are the means under control and treatment and `e` is the
propensity score of treatment.

We construct a confidence sequence imagining that the data was coming in
sequentially. If the

confidence sequence moves away from zero, we could have stopped
collecting data and deployed

whatever the treatment was. We also show the erroneous repeated
confidence intervals and if you run this repeated times you will see
some miscoverage events.

```{r}

library(ggplot2)
library(patchwork)

# --- Parameters ---
n <- 1e3; p0 <- 0.4; p1 <- 0.6; e <- 0.5
tau <- p1 - p0

running_sd <- function(x) {
  n <- seq_along(x)
  M2 <- numeric(length(x))
  mu <- cumsum(x) / n
  for (t in 2:length(x)) {
    d <- x[t] - mu[t - 1]
    M2[t] <- M2[t - 1] + d * (x[t] - mu[t])
  }
  sqrt(pmax(M2 / pmax(n - 1, 1), 1e-10))
}

robbins_confseq <- function(x, alpha = 0.05, rho = 1) {
  n      <- seq_along(x)
  mu_hat <- cumsum(x) / n
  s_n    <- running_sd(x)
  radius <- s_n * sqrt((2 * (n * rho^2 + 1) / (n^2 * rho^2)) *
                         log(sqrt(n * rho^2 + 1) / alpha))
  data.frame(lower = mu_hat - radius, upper = mu_hat + radius)
}

Z   <- rbinom(n, 1, e)
Y   <- ifelse(Z == 1, rbinom(n, 1, p1), rbinom(n, 1, p0))
phi <- Z * Y / e - (1 - Z) * Y / (1 - e)

cs  <- robbins_confseq(phi)

# CLT CI: just z * se, no correction for peeking
ns     <- 1:n
mu_hat <- cumsum(phi) / ns
clt_hw <- qnorm(0.975) * running_sd(phi) / sqrt(ns)

df <- data.frame(
  t        = ns,
  ate      = mu_hat,
  cs_lo    = cs$lower,   cs_hi = cs$upper,
  clt_lo   = mu_hat - clt_hw,
  clt_hi   = mu_hat + clt_hw
)[50:n, ]

p1_plot <- ggplot(df, aes(t)) +
  geom_ribbon(aes(ymin = cs_lo, ymax = cs_hi), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = ate), color = "steelblue") +
  geom_hline(yintercept = tau, color = "red", linetype = "dashed") +
  labs(x = "n", y = "ATE",
       title = "Asymptotic CS (valid)",
       subtitle = "Anytime-valid: accounts for peeking") +
  theme_minimal()

p2_plot <- ggplot(df, aes(t)) +
  geom_ribbon(aes(ymin = clt_lo, ymax = clt_hi), alpha = 0.2, fill = "coral") +
  geom_line(aes(y = ate), color = "coral") +
  geom_hline(yintercept = tau, color = "red", linetype = "dashed") +
  labs(x = "n", y = "ATE",
       title = "CLT CI (invalid for peeking)",
       subtitle = "Too narrow: misses true ATE when you peek") +
  theme_minimal()

p1_plot + p2_plot +
  plot_annotation(
    title    = sprintf("CS vs CLT CI | p0=%.1f, p1=%.1f, tau=%.1f", p0, p1, tau),
    subtitle = "Red dashed line = true ATE"
  )
```

## Miscellaneous

-   Group Sequential Designs are popular in FDA regulatory settings and
    in industry.

    -   The package `gsDesign` is the go-to package for this method

    -   Forces the analyst to pick times where they may stop the
        experiment

    -   My current research is related to these methods
